prolfqua - A Hany and Comprehensive R-package for Proteomics Differential Expression Analysis.

### 2022-05-31 inst/MyArticle/paper.Rmd
fix

We assessed the performance of different modeling approaches implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")`, `r BiocStyle::Biocpkg("MSstats")` and `r BiocStyle::Biocpkg("proDA")`, using a benchmark dataset whose ground truth is known. However, since only technical replicates are available for each dilution, essential sources of variation present in every experiment, such as the biochemical and biological, are not measured. The dataset captures only the variance from the chromatography, electro-spray, and mass spectrometric measurement method. Thus, while we can extrapolate some of the results obtained to more realistic datasets, we need to be careful not to over-interpret the results. Specifically, the observed variances will be higher in more realistic experiments, and sensitivities and specificity will be worse given the same sample sizes. Furthermore, the proportion of missing observations in real-life datasets might also be higher.

We can conclude that if we want to order the proteins according to the likelihood of being differentially regulated, to perform gene set enrichment analysis [@subramanian2005gene], the $t$-statistic is better suited than the fold-change estimate. Modeling the degrees of freedom when computing the $p$-values might further improve the inference, however, the observed this improvement is minuscule (see Figure \@ref(fig:FDRfdp) panel A).
However, there is no such effect for the mixed effect model, most likely because the degrees of freedom are erroneously estimated for a large proportion of models. Therefore, for the fixed effect linear model, the empirical Bayes variance shrinkage, as suggested by [@Smyth2004linear], consistently improves the ranking of proteins compared with the unmoderated estimates (not shown), however fails to do so for mixed effect models.

Computing the statistics at the peptide level, e.g., the $t$-statistics or $p$-value, then summarizing these statistics using their median produces the highest $AUC$ scores among all the tested models (see Figure \@ref(fig:FDRfdp) Panel A left). By using the beta distribution to model the number of peptides observed, we can further improve the $pAUC$ scores (see Figure \@ref(fig:FDRfdp) Panel A center). However, the properties of beta-based probabilities are not well understood; for instance, the p-values are not uniformely distributed under the null hypothesis (not shown). Furthermore, the FDR estimates obtained when correcting for multiple testing with the Benjamini-Hochberg method are biased overestimating the false discovery proportion (see Figure \@ref(fig:FDRfdp) Panel C). Therefore, we can not recommend this method if an unbiased estimate of FDR is essential, which is frequently the case. In addition, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset using this method (see Figure \@ref(fig:FDRfdp) Panel C).

>>>>

We assessed the performance of different modeling approaches implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")`, `r BiocStyle::Biocpkg("MSstats")` and `r BiocStyle::Biocpkg("proDA")`, using a benchmark dataset whose ground truth is known. However, since only technical replicates are available for each dilution, essential sources of variation present in every experiment, such as the biochemical and biological, are not measured. The dataset captures only the variance from the chromatography, electro-spray, and mass spectrometric measurement method. Thus, while we can extrapolate some of the results obtained to more realistic datasets, we need to be careful not to over-interpret the results. Specifically, the observed variances will be higher in more realistic experiments, and sensitivity and specificity will worsen, given the same sample sizes. Furthermore, the proportion of missing observations in real-life datasets might also be higher.

We can conclude that if we want to order the proteins according to the likelihood of being differentially regulated to perform gene set enrichment analysis [@subramanian2005gene], the $t$-statistic is better suited than the fold-change estimate. Modeling the degrees of freedom when computing the $p$-values might improve the inference. However, this improvement is minuscule (see Figure \@ref(fig:FDRfdp) panel A).
There is no such effect for the mixed effect model, most likely because the degrees of freedom are erroneously estimated for many models. Therefore, for the fixed effect linear model, the empirical Bayes variance shrinkage, as suggested by [@Smyth2004linear], consistently improves the ranking of proteins compared with the unmoderated estimates (not shown) and fails to do so for mixed effect models.

Computing the statistics at the peptide level, e.g., the $t$-statistics or $p$-value, then summarizing these statistics using their median produces the highest $AUC$ scores among all the tested models (see Figure \@ref(fig:FDRfdp) Panel A left). Furthermore, by using the beta distribution to model the number of peptides observed, we can further improve the $pAUC$ scores (see Figure \@ref(fig:FDRfdp) Panel A center). However, the properties of beta-based probabilities are not well understood; for instance, the p-values are not uniformly distributed under the null hypothesis (not shown). Furthermore, the FDR estimates obtained when correcting for multiple testing with the Benjamini-Hochberg method are biased and overestimate the false discovery proportion (see Figure \@ref(fig:FDRfdp) Panel C). Therefore, we can not recommend this method if an unbiased estimate of FDR is essential, which is frequently the case. In addition, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset using this method (see Figure \@ref(fig:FDRfdp) Panel C).
